---
title: "Accessing Human Cell Atlas Data"
author: "Maya Reed McDaniel"
date: "January 26th, 2021"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Accessing Human Cell Atlas Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```
# Motivation

- The purpose of this package is to make the querying of the
[HCA's Data Portal](https://www.humancellatlas.org/data-portal/) via their
provided
[API](https://data.humancellatlas.org/apis/api-documentation/data-browser-api)
easier and more accessible to researchers

## Human Cell Atlas (HCA) 

Visit the [Human Cell Atlas](https://data.humancellatlas.org/) for more
information on the project.

```{r install, eval = FALSE}
## install from Bioconductor if you haven't already
pkgs <- c("httr", "dplyr", "LoomExperiment", "hca")
pkgs_needed <- pkgs[!pkgs %in% rownames(installed.packages())]
BiocManager::install(pkgs_needed)
```

```{r setup, message = FALSE}
library(httr)
library(dplyr)
library(LoomExperiment)
library(hca)
```

# Example 1

- A useful workflow is to download and import the 'loom' file for the project
'Single cell transcriptome analysis of human pancreas reveals transcriptional
signatures of aging and somatic mutation patterns'.
- The project and loom file are in the 'dcp1' catalog.

1. Discover projects from the 'dcp1' (Data Coordinating Center, version 1)
catalog, with 'loom' files.

    ```{r}
    project_filter <- filters(fileFormat = list(is = "loom"))
    project_tibble <- projects(project_filter, catalog = "dcp1")
    project_tibble
    ```

2. Extracting the `projectId` for a project of interest
(`projectTitle` is "Single cell trans...")

    ```{r}
    project_tibble %>%
        filter(startsWith(projectTitle, "Single cell trans")) %>%
        t()
      
    projectId <-
        project_tibble %>%
        filter(startsWith(projectTitle, "Single cell trans")) %>%
        dplyr::pull(projectId)
    ```

3. Extract file information

    ```{r}
    file_filter <- filters(
        projectId = list(is = projectId),
        fileFormat = list(is = "loom")
    )
    
    file_tibble <- files(file_filter, catalog = "dcp1")
    ```
    
4. Download the loom file

    ```{r}
    file_tibble %>%
        files_download() %>%
        LoomExperiment::import()
    ```
    

# Example 2

This example illustrates access to `h5ad` files stored in the HCA and
created by the Python AnnData analysis software.

## Projects facets and terms

The first challenge is to understand what file formats are available
from the HCA. Obtain a tibble describing the 'facets' fo the data, the
number of terms used in each facet, and the number of distinct values
used to describe projects.

```{r}
projects_facets()
```

Note the `fileFormat` facet, and repeat `projects_facets()` to
discover detail about available file formats

```{r}
projects_facets("fileFormat")
```

Note that there are 8 uses of the `h5ad` file format. Use this as a
filter to discover relevant projects.

```{r}
filters <- filters(fileFormat = list(is = "h5ad"))
projects(filters)
```

## Projects columns

The default tibble produced by `projects()` contains only some of the
information available; the information is much richer.

### `projects()` as an _R_ `list`

Instead of retrieving the result of `projects()` as a tibble, retrieve
it as a 'list-of-lists'

```{r}
projects_list <- projects(as = "list")
```

This is a complicated structure. We will use `lengths()`, `names()`,
and standard _R_ list selection operations to navigate this a bit. At
the top level there rae three elements.

```{r}
lengths(projects_list)
```

`hits` represents each project as a list, e.g,.

```{r}
lengths(projects_list$hits[[1]])
```

shows that there are 10 different ways in which the project is
described. Each component is itself a list-of-lists, e.g.,

```{r}
lengths(projects_list$hits[[1]]$projects[[1]])
projects_list$hits[[1]]$projects[[1]]$projectTitle
```

### `projects()` as an `lol`

Use `as = "lol"` to create a more convenient way to select, filter and
extract elements from the list-of-lists by `projects()`.

```{r}
lol <- projects(as = "lol")
lol
```

Use `lol_select()` to restrict the `lol` to particular paths, and
`lol_filter()` to filter results to pathsleafs, or with specific
numbers of entries.

```{r}
lol_select(lol, "hits[*].projects[*]")
lol_select(lol, "hits[*].projects[*]") |>
    lol_filter(n == 44, is_leaf)
```

and the path or its abbreviation can be used to specify the columns of
the tibble to be returned by the `projects()` query.

### Creating `projects()` tibbles with specific columns

Here we retrieve additional details of donor count and total cells by
adding appropriate path abbreviations to a named character
vector. Names on the character vector can be used to rename the path
more consisely, but the paths must uniquely identify elements in the
list-of-lists.

```{r}
columns <- c(
    projectId = "hits[*].entryId",
    projectTitle = "hits[*].projects[*].projectTitle",
    genusSpecies = "hits[*].donorOrganisms[*].genusSpecies[*]",
    donorCount = "hits[*].donorOrganisms[*].donorCount",
    organ = "hits[*].cellSuspensions[*].organ[*]",
    totalCells = "hits[*].cellSuspensions[*].totalCells"
)
projects <- projects(filters, columns = columns)
projects
```

Note that the `organ` and `totalCells` columns have more than one
entry per project.

```{r}
projects |>
   select(projectId, organ, totalCells)
```

In this case, the mapping between `organ` and `totalCells` is clear,
but in general more refined navigation of the `lol` structure may be
necessary.

```{r}
projects |>
    select(projectId, organ, totalCells) |>
    filter(lengths(totalCells) > 0) |>
    tidyr::unnest(c("organ", "totalCells"))
```

Select the following entry, augement the filter, and query available files

```{r}
projects %>%
    filter(startsWith(projectTitle, "Reconstruct")) %>%
    t()
```

## File download

The relevant file can be selected and downloaded using the technique
in the first example.

```{r}
filters <- filters(
    projectId = list(is = "f83165c5-e2ea-4d15-a5cf-33f3550bffde"),
    fileFormat = list(is = "h5ad")
)
files <- 
    files(filters) %>%
    head(1)            # only first file, for demonstration
files %>% t()
```

```{r, eval = FALSE}
file_path <- files_download(files)
```

`"h5ad"` files can be read as SingleCellExperiment objects using the
[zellkonverter][] package.

```{r, eval = FALSE}
# don't want large amount of data written to disk
sce <- zellkonverter::readH5AD(file_path, use_hdf5 = TRUE)
# fails as matrix is transpose of expected matrix
sce
```

[zellkonverter]: https://bioconductor.org/packages/zellkonverter

# Example 3

- This provides an example of a multiple file download

```{r}
project_filter <- filters(fileFormat = list(is = "csv"))
project_tibble <- projects(project_filter, catalog = "dcp2")

project_tibble %>%
filter(startsWith(projectTitle,
                  "Reconstructing the human first trimester"))

projectId <- 
  project_tibble %>%
  filter(startsWith(projectTitle,
                    "Reconstructing the human first trimester")) %>%
  pull(projectId)

file_filter <- filters(
  projectId = list(is = projectId),
  fileFormat = list(is = "csv")
  )

## first 4 files will be returned
file_tibble <- files(file_filter, catalog = "dcp2", size = 4)

file_tibble %>%
  files_download()
```

# Example 4

This example explores the pagination feature of the results returned by
`projects()`, `files()`, `bundles()`, and `samples()`

- for the default tibble output
```{r}
page_1_tbl <- projects(size = 30)

page_2_tbl <- page_1_tbl %>% hca_next()

## should be identical to page_1_tbl
page_2_tbl %>% hca_prev()
```

- for the custom list-of-lists output
```{r}
page_1_lol <- projects(size = 30, as = "lol")

page_2_lol <- page_1_lol %>% hca_next()

## should be identical to page_1_lol
page_2_lol %>% hca_prev()
```