---
title: "Human Cell Atlas Manifests"
author: "Maya Reed McDaniel"
date: "September 2nd, 2021"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{uman Cell Atlas Manifests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

# Motivation & Introduction

The purpose of this vignette is to explore the file manifests available from the
[Human Cell Atlas](https://data.humancellatlas.org/) project.

These files provide a metadata summary for a collection of files in a tabular
format, including but not limited to information about process and workflow used
to generate the file, information about the specimens the file data were derived
from, and identifiers connect specific projects, files, and specimens.

The [WARP (WDL Analysis Research Pipelines) repository](https://broadinstitute.github.io/warp/docs/get-started)
contains information on a variety of pipelines, and can be used alongside a
manifest to better understand the metadata.

## Installation and getting started

Evaluate the following code chunk to install packages required for
this vignette.

```{r install, eval = FALSE}
## install from Bioconductor if you haven't already
pkgs <- c("dplyr", "LoomExperiment", "hca")
pkgs_needed <- pkgs[!pkgs %in% rownames(installed.packages())]
BiocManager::install(pkgs_needed)
```

Load the packages into your _R_ session.

```{r setup, message = FALSE}
library(dplyr)
library(LoomExperiment)
library(hca)
```

# Example: Default function parameters
```{r}
default_manifest_tbl <- hca::manifest()
default_manifest_tbl
```

# Example: Manifest for a specific catalog
```{r}
lungmap_manifest_tbl <- hca::manifest(catalog = "lungmap")
lungmap_manifest_tbl
```

# Example: Manifest for specific criteria

# Example: Using manifest data to select files

# Example: Using manifest data to annotate a `.loom` file
