---
title: "Accessing Human Cell Atlas Data"
author: "Maya Reed McDaniel"
date: "January 26th, 2021"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Accessing Human Cell Atlas Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```
# Human Cell Atlas

Visit the [Human Cell Atlas].

[Human Cell Atlas]: https://data.humancellatlas.org/

## Biological Background

- It is possible to isolate an individual cell and measure level of expression
of each gene in that cell
- The result is a matrix where:
    - rows = genes (~ N=30k)
    - columns = cells (~ M = 10k - 1mil)
    - matrix cells = expression values
- Experiments on the order of 10ks of cells and typically about 30k genes
- Matrices of expression values tend to be sparse i.e. 95% of cells may be zeros

## Experiment Workflow (pipeline)

1. Extract short DNA sequences from cells (millions per cell)
2. The short DNA sequences (about 100 nucleotide base pairs long) are analyzed
and output to FASTQ files
  - FASTQ files are often transformed to BAM or CRAM formats
  - BAM or CRAM files are summarized into the matrices
    - `.csv` file (standard)
    - `.loom` (based on `hdf5` *which is NOT hadoop*, but rather astrophysics
    derived file format for large data)
    - `.mtx` for sparse text-based matrix representation with 3 columns:
      1. row index
      2. column index, and
      3. non-zero value 

## HCA Projects

- HCA projects focus on a relatively small number of samples ex. 4 humans
- Each "project" is an experiment that is an entry in the overall database
    - some fields include the tissue organ type, species, etc.
    
## Motivation

- This R package was created to make the querying of the HCA's Data Portal via
their provided API easier and more accessible to researchers

```{r install, eval = FALSE}
## install from Bioconductor if you haven't already
pkgs <- c("httr", "dplyr", "LoomExperiment", "hca")
pkgs_needed <- pkgs[!pkgs %in% rownames(installed.packages())]
BiocManager::install(pkgs_needed)
```

```{r setup, message = FALSE}
library(httr)
library(dplyr)
library(LoomExperiment)
library(hca)
```

# Example 1

- A useful workflow is to download and import the 'loom' file for the project
'Single cell transcriptome analysis of human pancreas reveals transcriptional signatures of aging and somatic mutation patterns'.
- The project and loom file are in the 'dcp1' catalog.

1. Discover projects from the 'dcp1' (Data Coordinating Center, version 1) catalog, with 'loom' files.

    ```{r}
    project_filter <- filters(fileFormat = list(is = "loom"))
    project_tibble <- projects(project_filter, catalog = "dcp1")
    project_tibble
    ```

2. Extracting the `projectId` for a project I'm interested in
(`projectTitle` is "Single cell trans...")

    ```{r}
    project_tibble %>%
        filter(startsWith(projectTitle, "Single cell trans")) %>%
        t()
      
    projectId <-
        project_tibble %>%
        filter(startsWith(projectTitle, "Single cell trans")) %>%
        dplyr::pull(projectId)
    ```

3. File info

    ```{r}
    file_filter <- filters(
        projectId = list(is = projectId),
        fileFormat = list(is = "loom")
    )
    
    file_tibble <- files(file_filter, catalog = "dcp1")
    ```
    
4. Download the loom file(s)

    ```{r}
    file_tibble %>%
        files_download() %>%
        LoomExperiment::import()
    ```
    

# Example 2

This example illustrates access to `h5ad` files stored in the HCA and
created by the Python AnnData analysis software.

## Projects facets and terms

The first challenge is to understand what file formats are available
from the HCA. Obtain a tibble describing the 'facets' fo the data, the
number of terms used in each facet, and the number of distinct values
used to describe projects.

```{r}
projects_facets()
```

Note the `fileFormat` facet, and repeat `projects_facets()` to
discover detail about available file formats

```{r}
projects_facets("fileFormat")
```

Note that there are 8 uses of the `h5ad` file format. Use this as a
filter to discover relevant projects.

```{r}
filters <- filters(fileFormat = list(is = "h5ad"))
projects(filters)
```

## Projects columns

The default tibble produced by `projects()` contains only some of the
information available; the information is much richer.

### `projects()` as an _R_ `list`

Instead of retrieving the result of `projects()` as a tibble, retrieve
it as a 'list-of-lists'

```{r}
projects_list <- projects(as = "list")
```

This is a complicated structure. We will use `lengths()`, `names()`,
and standard _R_ list selection operations to navigate this a bit. At
the top level there rae three elements.

```{r}
lengths(projects_list)
```

`hits` represents each project as a list, e.g,.

```{r}
lengths(projects_list$hits[[1]])
```

shows that there are 10 different ways in which the project is
described. Each component is itself a list-of-lists, e.g.,

```{r}
lengths(projects_list$hits[[1]]$projects[[1]])
projects_list$hits[[1]]$projects[[1]]$projectTitle
```

### `projects()` as an `lol`

Use `as = "lol"` to create a more convenient way to select, filter and
extract elements from the list-of-lists by `projects()`.

```{r}
lol <- projects(as = "lol")
lol
```

Use `lol_select()` to restrict the `lol` to particular paths, and
`lol_filter()` to filter results to pathsleafs, or with specific
numbers of entries.

```{r}
lol_select(lol, "hits[*].projects[*]")
lol_select(lol, "hits[*].projects[*]") |>
    lol_filter(n == 44, is_leaf)
```

and the path or its abbreviation can be used to specify the columns of
the tibble to be returned by the `projects()` query.

### Creating `projects()` tibbles with specific columns

Here we retrieve additional details of donor count and total cells by
adding appropriate path abbreviations to a named character
vector. Names on the character vector can be used to rename the path
more consisely, but the paths must uniquely identify elements in the
list-of-lists.

```{r}
columns <- c(
    projectId = "hits[*].entryId",
    projectTitle = "hits[*].projects[*].projectTitle",
    genusSpecies = "hits[*].donorOrganisms[*].genusSpecies[*]",
    donorCount = "hits[*].donorOrganisms[*].donorCount",
    organ = "hits[*].cellSuspensions[*].organ[*]",
    totalCells = "hits[*].cellSuspensions[*].totalCells"
)
projects <- projects(filters, columns = columns)
projects
```

Note that the `organ` and `totalCells` columns have more than one
entry per project.

```{r}
projects |>
   select(projectId, organ, totalCells)
```

In this case, the mapping between `organ` and `totalCells` is clear,
but in general more refined navigation of the `lol` structure may be
necessary.

```{r}
projects |>
    select(projectId, organ, totalCells) |>
    filter(lengths(totalCells) > 0) |>
    tidyr::unnest(c("organ", "totalCells"))
```

Select the following entry, augement the filter, and query available files

```{r}
projects %>%
    filter(startsWith(projectTitle, "Reconstruct")) %>%
    t()
```

## File download

The relevant file can be selected and downloaded using the technique
in the first example.

```{r}
filters <- filters(
    projectId = list(is = "f83165c5-e2ea-4d15-a5cf-33f3550bffde"),
    fileFormat = list(is = "h5ad")
)
files <- files(filters)
files %>% t()
```

```{r, eval = FALSE}
file_path <- files_download(files)
```

`"h5ad"` files can be read as SingleCellExperiment objects using the
[zellkonverter][] package.

```{r, eval = FALSE}
# don't want large amount of data written to disk
sce <- zellkonverter::readH5AD(file_path, use_hdf5 = TRUE)
# fails as matrix is transpose of expected matrix
sce
```

[zellkonverter]: https://bioconductor.org/packages/zellkonverter

# Example 3

- This puts together the steps above, and provides an example of a case when
there is more than one file to download


<!-- ```{r} -->
<!--   project_filter <- filters(fileFormat = list(is = "bam")) -->
<!--   project_tibble <- projects(project_filter, catalog = "dcp1") -->

<!--   project_tibble %>% -->
<!--     filter(startsWith(projectTitle, "Structural Remodeling of the Human")) -->

<!--   projectId <- -->
<!--     project_tibble %>% -->
<!--     filter(startsWith(projectTitle, "Structural Remodeling of the Human")) %>% -->
<!--     pull(projectId) -->

<!--   file_filter <- filters( -->
<!--     projectId = list(is = projectId), -->
<!--     fileFormat = list(is = "bam") -->
<!--   ) -->
<!--   file_tibble <- files(file_filter, catalog = "dcp1", size = 2) -->

<!--   ## takes about an hour to run -->
<!--   file_tibble %>% -->
<!--     download_files() -->
<!-- ``` -->

